Notes

### 2022-09-22 notes/notes.tex
add discussion about optimality

To balance these two criteria, we measure the quality of the model as 
\begin{equation}
    \label{eq: cost function regularised}
f(W) := 
\underbrace{\frac{1}{2}\norm{\log h.(\hat{y})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{\log W}^2}_{\text{deviation from average emission}}
\quad\text{where}\quad 
\hat{y} = M_{shift} W \tilde{x}
\end{equation}
where $\lambda \ge 0$ is a regularisation parameter.

To optimise $f$ numerically, we will parametrise $W$ as $W = e^{\mathrm{diag}\, v}$ where $v = [v_1, v_2,\dots]$ is some vector and $e^{\mathrm{diag}\, v}$ is the diagonal matrix with $[e^{v_1}, e^{v_2},\dots]$ on the diagonal. The motivation for doing this is twofold:
\begin{enumerate}
    \item It ensures that $W$ remains strictly positive. In particular, the diagonal elements of $W$ can only tend towards zero if $v_j \rightarrow -\infty$ for some $i$. 
    \item It implicitly measures changes to $W$ in a relative sense. If the optimisation algorithm performs a linear update $v_{k+1} = v_k + \Delta v$, this corresponds to a relative update $W_{k+1} = W_k e^{\mathrm{diag}\, \Delta v}$.
\end{enumerate}

Thus, in practice, we are interested in the optimisation problem
$$
\min_{v} f(v) =
\frac{1}{2}\norm{\log h.(\hat{y})}^2 + \frac{1}{2}\lambda \norm{v}^2
\quad\text{where}\quad 
\hat{y} = M_{shift} e^{\mathrm{diag}\, v} \tilde{x}
.$$
Most optimisation algorithms require at least the gradient of $f$. This is given by the following:
\begin{proposition}
    Let $g = [g_1, \dots, g_n]$ be the vector defined elementwise by 
    $$
    g_i = \frac{\log h(\hat{y}_i)}{h(\hat{y}_i)} \frac{\mathrm{d}}{\mathrm{d}\hat{y}_i} h(\hat{y}_i)
    .$$
    Then 
    $$
    \mathrm{grad}\, f(v) = e^v \odot (M_{shift}^T g) \odot \tilde{x} 
    $$
    where $e^v$ denotes elementwise exponentiation and $\odot$ is the Hadamard (elementwise) product.
\end{proposition}
\begin{proof}

>>>>

To balance these two criteria, we measure the quality of the model as 
\begin{equation}
    \label{eq: cost function regularised}
f(W) := 
\underbrace{\frac{1}{2}\norm{\log h.(\hat{y}(W))}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{\log W}^2}_{\text{deviation from average emission}}
\quad\text{where}\quad 
\hat{y}(W) = M_{shift} W \tilde{x}
\end{equation}
where $\lambda \ge 0$ is a regularisation parameter.

Observe that the model where $W$ is the identity matrix $\mathds{I}$ is equivalent to the original model. Furthermore, the regularisation term vanishes if and only if $W = \mathds{I}$. That is,
$$
f(\mathds{I}) = \frac{1}{2}\norm{\log h.(\hat{y}(\mathds{I}))}^2
$$
is the prediction error for the original model.
For $W \ne \mathds{I}$, the regularisation term is strictly positive. This means that, if $f(W) \le f(\mathds{I})$, then the prediction error is decreased, i.e.,
$$
\frac{1}{2}\norm{\log h.(\hat{y}(W))}^2
\le
\frac{1}{2}\norm{\log h.(\hat{y}(\mathds{I}))}^2
.$$
Hence, the model corresponding to the minimum of \cref{eq: cost function regularised} predicts the data at least as accurately as the original model.

To optimise $f$ numerically, we will parametrise $W$ in \cref{eq: cost function regularised} as $W = e^{\mathrm{diag}\, v}$ where $v = [v_1, v_2,\dots]$ is some vector and $e^{\mathrm{diag}\, v}$ is the diagonal matrix with $[e^{v_1}, e^{v_2},\dots]$ on the diagonal. The motivation for doing this is twofold:
\begin{enumerate}
    \item It ensures that $W$ remains strictly positive. In particular, the diagonal elements of $W$ can only tend towards zero if $v_j \rightarrow -\infty$ for some $i$. 
    \item It implicitly measures changes to $W$ in a relative sense. If the optimisation algorithm performs a linear update $v_{k+1} = v_k + \Delta v$, this corresponds to a relative update $W_{k+1} = W_k e^{\mathrm{diag}\, \Delta v}$.
\end{enumerate}

Thus, in practice, we are interested in the optimisation problem
$$
\min_{v} f(v) =
\frac{1}{2}\norm{\log h.(\hat{y})}^2 + \frac{1}{2}\lambda \norm{v}^2
\quad\text{where}\quad 
\hat{y} = M_{shift} e^{\mathrm{diag}\, v} \tilde{x}
.$$

Most optimisation algorithms require at least the gradient of $f$. This is given by the following:
\begin{proposition}
    Let $g = [g_1, \dots, g_n]$ be the vector defined elementwise by 
    $$
    g_i = \frac{\log h(\hat{y}_i)}{h(\hat{y}_i)} \frac{\mathrm{d}}{\mathrm{d}\hat{y}_i} h(\hat{y}_i)
    .$$
    Then 
    $$
    \mathrm{grad}\, f(v) = e^v \odot (M_{shift}^T g) \odot \tilde{x} 
    $$
    where $e^v$ denotes elementwise exponentiation and $\odot$ is the Hadamard (elementwise) product.
\end{proposition}
\begin{proof}
